Future Cosmic Microwave Background delensing with galaxy surveys

### 2017-10-04 Notes/galaxies_delensing_uchicago_thesis.tex
Kimmy and Toshiya input

%% Use these commands to set biographic information for the title page:
\title{Future Cosmic Microwave Background delensing with galaxies surveys}
\author{Alessandro Manzotti}
\department{Astronomy and Astrophysics}
\division{Physical Sciences}
\degree{Doctor of Phylosophy}
\date{August 2017}
\dedication{Dedication Text}
\epigraph{No epigraph}

@@

\maketitle
\makecopyright
%\makededication
%\makeepigraph
\tableofcontents
\listoffigures
\listoftables

@@

The cosmic microwave background (CMB) polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. 
A particular component, the so-called B-modes, is indeed a direct signature of a prediction of inflation: the presence of gravitational waves in the early universe. However, improving the level of noise in future experiments will not be enough. 
Secondary effects in the low redshift universe will also produce non-primordial B-modes adding confusion to the inflationary signal. In particular, the gravitational interactions of CMB photons with large scale structures will distort the primordial E-modes adding a lensing B-mode component to the primordial signal.
Removing the lensing component from the measurement of CMB B-modes will then be necessary to constrain the amplitude of the primordial gravitational waves.
Here we discuss the role of current and future large scale structure surveys in multi-tracers approach that will improve the reconstruction of the lensing potential that lenses the CMB photons and and, as a consequence, the delensing efficiency. 
We quantify this by the improvement due to delensing on the constraints on the inflationary tensor perturbations amplitude and shape ($r$ and $n_{t}$).
We find that a galaxy survey like DES can remove about $14\%$ of the lensing signal. current generation CMB experiments will benefit from 
With the improvement of the noise CMB internal reconstruction will become more important and the fraction of removed lensing B-modes will rapidly improve from the current level of Planck ($8\%$) and SPTPol ($35\%$) to 3G (56\%) and CMB S4 (76\%) level. 
Nevertheless optical galaxy surveys, in particular if split in different tomographic bin will still play an important role even for CMB S4. 
\mainmatter

\chapter{Introduction}
\label{sec:intro}
In the standard cosmological paradigm the early universe underwent a period of near-exponential expansion called "cosmic inflation." 
All the cosmological observations agree with this picture, making it a compelling and elegant description of the Universe initial conditions. Despite the experimental effort, other possible explanations are still valid, and a conclusive evidence of inflation is still to be found. 
Inflation generically predicts a stochastic background of gravitational waves \cite[see e.g.][for a review]{kamionkowski15}. This prediction set inflation apart from other theories and a detection of primordial gravitational waves could be the compelling evidence cosmologists are looking for. 
These primordial gravitational waves in the early universe would imprint a unique signature on the polarized anisotropies of the CMB. For this reason, CMB polarization is a promising dataset to understand the physics of the early universe and ultimately test inflation.
In particular, we can decompose the CMB polarization fields in Fourier space into even-parity (divergence) and odd-parity (curl) components, referred to as ``E'' and ``B'' modes.
 In the standard scenario, the B-mode polarization is a clean probe of primordial gravitational waves, because these are the only source of B-modes at the epoch of recombination.

Because the CMB B-modes provide the cleanest known observational window into the primordial gravitational waves background, improving their measurement is a major objective of current and future CMB experiments.
Even if inflationary B-modes have not been detected yet, a natural value $r \gtrsim 10^{-3}$ should be reachable shortly given the level of noise expected in future CMB experiments \citep{kamionkowski15}.
However, just reducing the level of noise will not be enough to attain this goal.
Indeed, the observed B modes are not solely sourced by early universe physics; they are also produced by secondary effects taking place in the late-time low-redshift universe. 
In this work, we will focus on the effect of gravitational interactions with large scale structures (LSS).
Lensing shears the CMB polarization pattern, producing ``lensing B modes'' from CMB E modes \citep{zaldarriaga98}.
This expected component has now been measured both in cross-correlation with LSS \citep{hanson13,polarbear2014c,vanengelen14b,planck2015XV} and from CMB data alone \citep{polarbear2014b,bicep2a,keisler15}.

This component acts as a source of confusion for searches of the primordial gravitational wave background.
Indeed, the contamination from lensing B modes is already at the level of the instrumental noise of current experiments \citep{bk14}. Thus, together with the experimental effort to reduce the amount of noise, the effect of the spurious lensing component must be understood.

The optimal way to reduce the lensing contributions to the B-modes is to reconstruct the expected realization of lensing B modes on the observed part of the sky and then use it to clean the data in a process called ``delensing''. 
We can delens the observed B-modes by combining CMB polarization data (what is \textit{lensed}) with tracers of the large scale structure (what is \textit{lensing}) to reconstruct a template of the expected lensing B-modes.
Delensing has been studied for many years \cite{knox2002, kesden2002, seljak2003, simard:2015,sherwin15,smith:2012}.
Furthermore, it has recently been performed on CMB temperature data using the cosmic infrared background as LSS tracer \cite{larsen:2016} and on CMB temperature and polarization data using CMB data to internally reconstruct the LSS lensing potential \cite{ carron17}. Finally, the highest B-mode delensing efficiency has been achieved with SPT and Herschel data in which $~28\%$ of the lensing power was removed.

@@

Large scale structure surveys usually probe the 3D dark matter overdensities as a 2D field projected along the line of sight:

\begin{equation}
\delta^{i}(\nver) = \int_0^{\infty} dz\, W^{i}(z)\delta(\chi(z)\nver,z).
\label{eqn:wkernel}
\end{equation}

where $\delta(\chi(z)\nver,z)$ corresponds to the dark matter overdensity field at a comoving distance $\chi(z)$ and at a redshift $z$ in the angular direction $\nver$.
Using the Limber approximation \cite{limber53} we can compute the power spectra of two large-scale structure fields $i,j$ as:
\begin{equation}
C_{\ell}^{ij}= \int_0^{\infty} \frac{dz}{c} \frac{H(z)}{\chi(z)^2}\, W^{i}(z)W^{j}(z)P(k,z).
\label{eqn:wkappa}{}
\end{equation}

In this equation, $H(z)$ is the Hubble factor at redshift $z$, $c$ is the speed of light and $P(k,z)$ is the matter power spectrum evaluated at wavenumber $k = \ell/\chi(z)$ and redshift $z$. 
Furthermore $W^{i}(z)$ is the kernel function of the field $i$ introduced in \refeq{wkernel}.
We will now describe the kernels for each of the tracers used in this work.

\section{CMB lensing potential}
\label{sec:kappaCMB}
We start from the CMB lensing potential.
The lensing kernel $W^{\kappa}$ is:
\begin{equation}
W^{\kappa}(z) = \frac{3\Omega_{\rm m}}{2c}\frac{H_0^2}{H(z)}(1+z)\chi(z)\frac{\chi_*-\chi(z)}{\chi_*},
\end{equation}
%
where $\chi_*$ is the comoving distance to the last-scattering surface at $z_*\simeq 1090$, $\Omega_{\rm m}$ and $H_0$ are the present day values of the Hubble and matter density parameters, respectively.

The CMB lensing potential is \textit{the} field that we need to reconstruct to reverse the effect of large scale structure and delens the CMB.
However, the lensing potential can also be reconstructed using the CMB itself. In that case, we can treat it as a noisy tracer of the true field. Both the CMB lensing field and its noisy reconstructed counterpart have the same kernel $W^{\kappa}(z)$.
However, when computing the power spectrum of the latter, we need to add a noise component. Given the instrumental noise level and the beam, we can calculate the reconstruction noise $N_{\ell}^{\kappa \kappa}$, and so
\beq\label{eqn:noisekappa}
\cl^{\kappa_\mathrm{rec} \kappa_\mathrm{rec}} = \cl^{\kappa \kappa } +N^{\kappa \kappa}_{\ell}
\eeq
In this work, the level of noise is computed assuming an iterative approach to the CMB lensing reconstruction as described in \cite{smith:2012,hirata:2003}.

\section{Galaxies}

The galaxy clustering kernel is:
\begin{equation}
\label{eqn:wg}
\begin{split}
W^{g}(z) &= \frac{b(z)\frac{dN}{dz}}{\Bigl(\int dz'\,\frac{dN}{dz'}\Bigr)}.
\end{split}
\end{equation}
Here $\frac{dN}{dz}$ is the number of galaxies observed by the survey as a function of redshift while $b(z)$ is the galaxy bias that connects the amplitude of galaxy overdensities to the underlying dark matter density.
When computing the auto-spectrum of the galaxy density, a shot noise term needs to be taken into account. 
To do so, we add a constant term to the power spectrum equal to the inverse of the number of galaxies per steradians.
 Different galaxy surveys in this work are then fully characterized by their b(z), $\frac{dN}{dz}$ and the observed galaxy density.
We test the delensing efficiency taking into account both current surveys like WISE or DES as well as future galaxy surveys like DESI and LSST together with 21 cm measurement like SKA.

@@

DES is modeled after the DES Science Verification public data release.
For DESI we used the $\frac{dN}{dz}$ in Tab. 2.3 of the DESI Technical Design Report. From that we can derive the galaxy denisty of 0.63 galaxies per arcmin squared. 
%As in \cite{passaglia:2017} we use a linear bias $b$

For LSST we follow \cite{schaan:2016}: $\frac{dN}{dz}\propto z^{\alpha}\exp^{-(z/z_{0})^{\beta}}$ with $\alpha = 1.27$, $\beta = 1.02$, and $z_{0}= 0.5$. Furthermore we assume a density of 26 galaxies per arcmin squared.

Finally, we consider the Square Kilometre Array (SKA).
The SKA is a planned radio array that will survey large scale structure primarily by detecting the redshifted neutral hydrogen (HI) 21cm emission line from a large number of galaxies out to high redshift. 
We will assume an intensity mapping survey mode where SKA will measure the large-scale fluctuations of the \textit{integrated} 21cm intensity from many unresolved galaxies.  
We model both the redshift distribution and  bias of radio sources following \cite{namikawa:2016a}.

@@

\section{Cosmic infrared Background (CIB)}
\label{sec:cib}

@@

Following \cite{Addison:2011se}, we model the CIB power directly as $C_{\ell}^{{\rm CIB}\mbox{-}{\rm CIB}} = 3500 (l/3000)^{-1.25} {\rm Jy^2 / sr}$.
This model provides an accurate fit to several experimental results.
For the cross-spectra with the CMB lensing or other galaxy tracers, $C_{\ell}^{{\rm CIB}\mbox{-}j}$, we use the single-SED model of~\cite{hall10}.
It corresponds to the kernel:
\begin{align}\label{eq:12}
W^{\text{CIB}}(z) = b_c\ \frac{\chi^2(z)}{H(z)(1+z)^2}\ e^{-\frac{(z-z_c)^2}{2\sigma^2_z} } f_{\nu(1+z)},
\end {align}
for
\begin{equation}
f_{\nu} = 
\begin{cases}
\Big( e^{\frac{h\nu}{kT}} - 1 \Big)^{-1} \nu^{\beta+3} & (\nu \leq v^{\prime}) \\ \Big( e^{\frac{h\nu^{\prime}}{kT}} - 1 \Big)^{-1} \nu^{\prime \beta+3} \Big( \frac{\nu}{\nu^{\prime}} \Big)^{-\alpha} & (\nu > v^{\prime})
\end{cases}
\end{equation}

We place the peak of the CIB emissivity at redshift $z_c = 2$ with a broad redshift kernel of width $\sigma_z = 2$ and we set $T$ = 34K and $\nu^{\prime} \approx$ 4955 GHz.

\chapter{Gravitational lensing B-mode and delensing}
\label{sec:th}
The large scale structures described in \refsec{model} have an important impact on the search of primordial CMB B-modes: they lens the primordial E-modes generating non inflationary B-modes that constitues an important source of noise.

Indeed the Q and U mode decompositions of the CMB photons polarization are remapped by lensing as:
\be
Q(\hat{\mathbf{n}}) = Q_{\mathrm{unlensed}}(\hat{\mathbf{n}}+\nabla\phi);~~
U(\hat{\mathbf{n}}) = U_{\mathrm{unlensed}}(\hat{\mathbf{n}}+\nabla\phi)
\ee
where the deflection angle is the gradient of the lensing potential integrated along the line of sight $\nabla\phi$.
The CMB polarization is usually decomposed into odd-parity Fourier modes E and B. 
As shown in \cite{}, because of the symmetry of the problem, tensor perturbations are the principal source of the B-modes configuration. For this reason, B-modes are a promising signature of early universe tensor perturbations.

However, this promising sign of primordial gravitational waves is partially obscured by gravitational interactions with large scale structures that generate CMB B-modes by distorting primordial E-modes.
At first order, given the convergence field $\kappa= -\frac{1}{2}\nabla^2\phi$ introduced in \refsec{kappaCMB} the B-modes resulting from the lensing of primordial E-modes are:
\be
B^{\mathrm{lens}}(\bl) =  \int \frac{d^2 \bl'}{(2 \pi)^2} W(\bl,\bl') E(\ublu) \kappa(\bl - \bl')
\label{eqn:blens}
\ee
where different modes contributes with a different weight:
\be
W(\bl,\bl') = \frac{2 \bl' \cdot (\bl-\bl')}{|\bl-\bl'|^2} \sin(2\varphi_{\bl,\bl'}).
\ee
%As usual we define the power spectrum as:
%\be
%\langle B^{\mathrm{lens}}(\bl) ~ B^{\mathrm{lens}^*}(\tilde{\bl}) \rangle \equiv (2\pi)^2 \delta^D(\bl - \tilde{\bl}) C^{BB,\mathrm{lens}}_l .
%\ee

From this we get the power spectrum of the lensing component of the B-modes:
\be
C_{\ell}^{BB,\mathrm{lens}}  = \int \frac{d^2 \bl'}{(2 \pi)^2} W^2 (\bl,\bl')C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}.
\ee

The B-mode power spectrum measured on the sky is composed of a possible primordial component $C_{\ell}^{BB,r}$ together with the lensing $C_{\ell}^{BB,\mathrm{lens}}$ contribution and the instrumental noise $N_{\ell}^{BB}$ (defined in \refeq{noise}):
\be
C_{\ell}^{BB,\mathrm{full}} = C_{\ell}^{BB,r} + C_{\ell}^{BB,\mathrm{lens}} + N_{\ell}^{BB}.
\ee
The lensing component is a significant source of B-modes that, at large scales, correspond to a white noise source of roughly $5 \mu K$-arcmin. This means that it is not only bigger than the allowed inflationary component at scales smaller than several degrees, but it is also comparable with current levels of instrumental noise. 
For this reason, it is critical to characterize and eventually remove it from the data.
To do so, we build a template \refeq{blens} of the lensing B-modes in the observed patch given a measurement of the E-mode field and the lensing potential $\phi$.
While E is measured directly, we can estimate $\phi$ using "tracers" of the dark matter distribution that sources the potential.

We will now show how the delensing efficiency is related to the fidelity of the lensing tracers and the instrumental noise in the CMB E-modes.
If we have a large scale structure field I($\nver$) that traces the lensing potential we can build a template of the lensing B-modes on the sky by a weighted convolution:
\be
\hat{B}^{\mathrm{lens}}(\bl) = \int \frac{d^2 \bl'}{(2 \pi)^2} W(\bl,\bl') f(\bl,\bl') E^N(\ublu) I(\bl - \bl'),
\ee
where $f(\bl,\bl')$ can be determined by minimizing the difference with the true $B^{\mathrm{lens}}(\bl)$ defined in \refeq{blens}. We introduce the instrumental noise in the CMB E-modes ($E^N$) that will also limit the ability to fully reconstruct the lensing B-modes.

@@

The optimal weights $f(\bl,\bl')$ such that the residual lensing B mode power is minimized are:
\be
\label{eqn:fweight}
f(\bl,\bl') = \left(\frac{C^{EE}_{{l'}}}{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right)  \frac{C^{\kappa I}_{|\bl-\bl'|}}{C^{II}_{|\bl-\bl'|} }.
\ee
Here $C^{\kappa I}$ and $C^{II}$ are the cross-correlation spectrum of the tracer $I$ with the lensing convergence $\kappa$ and its autospectrum; they are described for each LSS field in \refsec{model}. The power spectrum of the E-modes noise $N^{EE}$ is the same as the B-modes one in \refeq{noise}.

Notice that the first term consists of an inverse variance filter applied to the measured E-mode and the second minimizes the difference between the reconstructed $\phi$ and the CMB lensing potential.

With this choice of $f(\bl,\bl')$ we finally have that the residual power is:
\beqn
\label{eqn:Bres}
C_{{l}}^{BB,\mathrm{res}} &=&   \int \frac{d^2 \bl'}{(2 \pi)^2}  W^2 (\bl,\bl')
 C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}  \\
&\times&  \left[1 - \left(\frac{C^{EE}_{{l'}} }{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right) \rho^2_{|\bl-\bl'|} \right] \nonumber
\eeqn
with
\be\label{eqn:rho1}
\rho_{\ell}^{2}= \frac{(\cl^{\kappa I})^{2}}{\cl^{\kappa \kappa}\cl^{I I}}.
\ee

>>>>

%% Use these commands to set biographic information for the title page:
\title{Future Cosmic Microwave Background delensing with galaxy surveys}
\author{Alessandro Manzotti}
\department{Astronomy and Astrophysics}
\division{Physical Sciences}
\degree{Doctor of Philosophy}
\date{August 2017}
\dedication{Dedication Text}
\epigraph{No epigraph}

@@

\maketitle
\makecopyright
\tableofcontents
\listoffigures
\listoftables

@@

The cosmic microwave background (CMB) polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. 
A particular component, the so-called B-modes, is indeed a direct signature of a prediction of inflation: the presence of gravitational waves in the early universe. However, reducing the instrumental noise in future experiments will not be enough to detect this signal. 
Secondary effects in the low redshift universe will also produce non-primordial B-modes adding confusion to the inflationary signal. In particular, the gravitational interactions of CMB photons with large scale structures will distort the primordial E-modes adding a lensing B-mode component to the primordial signal.
Removing the lensing part (``delensing'') from the measurement of CMB B-modes will then be necessary to constrain the amplitude of the primordial gravitational waves.
Here we discuss the role of current and future large scale structure surveys in a multi-tracers approach to delensing that will improve the reconstruction of the lensing potential that lenses the CMB photons and, as a consequence, the delensing efficiency. 
We quantify this by the improvement due to delensing on the constraints on the inflationary tensor perturbations amplitude and shape ($r$ and $n_{t}$). We find that, in general, galaxy surveys should be split into tomographic bins as this can improve the correlation with CMB lensing by 30\%. 
Among currently available surveys, a DES-like galaxy survey can remove about $14\%$ of the lensing signal. Ongoing CMB experiments (CMB-S2) will particularly benefit from large scale structure tracers that, once properly combined, will have a better performance than a CMB internal reconstruction.
With the decrease of instrumental noise, the CMB internal reconstruction will increase its efficiency and the fraction of removed lensing B-modes with CMB alone will rapidly improve from the current level of Planck ($8\%$) and SPTPol ($35\%$) to 3G (56\%) and CMB S4 (85\%) level.  
Nevertheless optical galaxy surveys will still play an important role even for CMB S4. In particular an LSST-like survey can a achieve a delensing performance comparable to a 3G CMB experiment but with completely different systematics. This will be important to prove the robustness against systematics of an eventual detection of primordial B-modes.
% maybe non linear scales?

\vspace{1cm}
\mainmatter
\chapter{Introduction}
\label{sec:intro}
In the standard cosmological paradigm the early universe underwent a period of near-exponential expansion called ``cosmic inflation''.
All the cosmological observations agree with this picture, making it a compelling and elegant description of the Universe initial conditions. Despite the experimental effort, other possible explanations are still valid, and a conclusive evidence of inflation is still to be found. 
Inflation generically predicts a stochastic background of gravitational waves \cite[see e.g.][for a review]{kamionkowski:2016}. This prediction sets inflation apart from other theories and a detection of primordial gravitational waves could be the compelling evidence cosmologists are looking for. 
These primordial gravitational waves in the early universe would imprint a unique signature on the polarized anisotropies of the CMB. For this reason, CMB polarization is a promising dataset to understand the physics of the early universe and ultimately test inflation.
In particular, we can decompose the CMB polarization fields in Fourier space into even-parity (divergence) and odd-parity (curl) components, referred to as ``E'' and ``B'' modes.
 In the standard scenario, the B-mode polarization is a clean probe of primordial gravitational waves, because these are the only source of B-modes at the epoch of recombination.

Because the CMB B-modes provide the cleanest known observational window into the primordial gravitational wave background, improving their measurement is a major objective of current and future CMB experiments.
Even if inflationary B-modes have not been detected yet, the target value $r \gtrsim 10^{-3}$, should be reachable shortly given the level of noise expected in future CMB experiments \citep{kamionkowski:2016,abazajian:2016}.
However, just reducing the level of noise will not be enough to attain this goal.
Indeed, the observed B modes are not solely sourced by early universe physics; they are also produced by secondary effects taking place in the late-time low-redshift universe. 
In this work, we will focus on the effect of gravitational interactions with large scale structures (LSS).
Lensing shears the CMB polarization pattern, producing ``lensing B modes'' from CMB E modes \citep{zaldarriaga98}.
This expected component has now been measured both in cross-correlation with LSS \citep{hanson13,bicep2-collaboration:2016,polarbear2014c,van-engelen:2015,planck2015XV} and from CMB data alone \citep{polarbear2014b,bicep2a,keisler15}.

This component acts as a source of confusion for searches of the primordial gravitational wave background.
Indeed, the contamination from lensing B modes is already at the level of the instrumental noise of current experiments \citep{bk14}. Thus, together with the experimental effort to reduce the amount of noise, the effect of the lensing component must be understood.

The optimal way to reduce the lensing contributions to the B-modes is to reconstruct a template of the actual lensing B modes on the observed part of the sky and then use it to clean the data in a process called ``delensing''. 
We can delens the observed B-modes by combining CMB polarization data (what is \textit{lensed}) with tracers of the large scale structure (what is \textit{lensing}) to reconstruct a template of the expected lensing B-modes.
Delensing has been studied for many years \cite{knox2002, kesden2002, seljak2003, simard:2015,sherwin15,smith:2012}.
Furthermore, it has recently been performed on CMB temperature data using the cosmic infrared background (CIB) as LSS tracer \cite{larsen:2016a} and on CMB temperature and polarization data using CMB data to internally reconstruct the LSS lensing potential \cite{carron:2017}. Finally, the highest B-mode delensing efficiency has been achieved with SPT and Herschel data in which $~28\%$ of the lensing power was removed \cite{manzotti:2017}.

@@

Large scale structure surveys usually probe the 3D matter overdensities as a 2D field projected along the line of sight:
\begin{equation}
\delta^{i}(\nver) = \int_0^{\infty} dz\, W^{i}(z)\delta(\chi(z)\nver,z).
\label{eqn:wkernel}
\end{equation}
where $\delta(\chi(z)\nver,z)$ corresponds to the dark matter overdensity field at a comoving distance $\chi(z)$ and at a redshift $z$ in the angular direction $\nver$.
Using the Limber approximation \cite{limber53} we can compute the power spectra of two large-scale structure fields $i,j$ as:
\begin{equation}
C_{\ell}^{ij}= \int_0^{\infty} \frac{dz}{c} \frac{H(z)}{\chi(z)^2}\, W^{i}(z)W^{j}(z)P(k,z).
\label{eqn:wkappa}{}
\end{equation}

In this equation, $H(z)$ is the Hubble factor at redshift $z$, $c$ is the speed of light, and $P(k,z)$ is the matter power spectrum evaluated at wavenumber $k = \ell/\chi(z)$ and redshift $z$. 
%Furthermore $W^{i}(z)$ is the kernel function of the field $i$ introduced in \refeq{wkernel}.
We will now describe the kernels $W^{i}(z)$ for each of the tracers used in this work.

\subsection{CMB lensing potential}
\label{sec:kappaCMB}
We start from the CMB lensing potential.
The lensing kernel $W^{\kappa}$ is:
\begin{equation}
W^{\kappa}(z) = \frac{3\Omega_{\rm m}}{2c}\frac{H_0^2}{H(z)}(1+z)\chi(z)\frac{\chi_*-\chi(z)}{\chi_*},
\end{equation}
%
where $\chi_*$ is the comoving distance to the last-scattering surface at $z_*\simeq 1090$, $\Omega_{\rm m}$ and $H_0$ are the present day values of the Hubble and matter density parameters, respectively. As can be seen in \reffig{kernel}, the lensing kernel $W^{\kappa}(z)$ is a broad function of redshift peaking around $z=1.8$ but extending to high redshift.

The CMB lensing potential is \textit{the} field that we need in order to reverse the effect of large scale structure and delens the CMB.
However, the lensing potential can also be reconstructed using the CMB itself. In that case, we can treat it as a noisy tracer of the true field. Both the CMB lensing field and its noisy reconstructed counterpart have the same kernel $W^{\kappa}(z)$.
However, when computing the power spectrum of the latter, we need to add a noise component. Given the instrumental noise level and the beam, we can calculate the reconstruction noise $N_{\ell}^{\kappa \kappa}$, and so
\beq\label{eqn:noisekappa}
\cl^{\kappa_\mathrm{rec} \kappa_\mathrm{rec}} = \cl^{\kappa \kappa } +N^{\kappa \kappa}_{\ell}
\eeq
In this work, the level of noise is computed assuming an iterative approach to the CMB lensing reconstruction as described in \cite{smith:2012,hirata:2003}.

\subsection{Galaxies}

The galaxy clustering kernel is:
\begin{equation}
\label{eqn:wg}
\begin{split}
W^{g}(z) &= \frac{b(z)\frac{dN}{dz}}{\Bigl(\int dz'\,\frac{dN}{dz'}\Bigr)}.
\end{split}
\end{equation}
Here $\frac{dN}{dz}$ is the number of galaxies observed by the survey as a function of redshift while $b(z)$ is the galaxy bias that connects the amplitude of galaxy overdensities to the underlying dark matter density. 
We use a linear bias independent of the angular scale, which is a reasonable assumption for the relatively large scales relevant for delensing ($\ell<1000$).
When computing the auto-spectrum of the galaxy density, a shot noise term needs to be taken into account. 
To do so, we add a constant term to the power spectrum equal to the inverse of the number of galaxies per steradians.
 Different galaxy surveys in this work are then fully characterized by their b(z), $\frac{dN}{dz}$ and the observed galaxy density.
We test the delensing efficiency taking into account both current surveys like WISE or DES as well as future galaxy surveys like DESI and LSST as well as 21 cm measurement like SKA.

@@

DES is modeled after the DES Science Verification public data release.
For DESI we used the $\frac{dN}{dz}$ in Tab. 2.3 of the DESI Technical Design Report. From that we can derive the galaxy denisty of 0.63 galaxies per squared arcmin. 
%As in \cite{passaglia:2017} we use a linear bias $b$

For LSST we follow \cite{schaan:2017}: $\frac{dN}{dz}\propto z^{\alpha}\exp^{-(z/z_{0})^{\beta}}$ with $\alpha = 1.27$, $\beta = 1.02$, and $z_{0}= 0.5$. Furthermore we assume a density of 26 galaxies per arcmin squared.

Finally, we consider the Square Kilometre Array (SKA).
The SKA is a planned radio array that will survey large scale structure primarily by detecting the redshifted neutral hydrogen (HI) 21cm emission line from a large number of galaxies out to high redshift. 
We will assume an intensity mapping survey mode where SKA will measure the large-scale fluctuations of the \textit{integrated} 21cm intensity from many unresolved galaxies.  
We model both the redshift distribution and  bias of radio sources following \cite{namikawa:2016a}. 
In \reffig{kernel}, we compare the CMB lensing kernel $W^{\kappa}(z)$ to the kernels $W^{g}(z)$ of all the tracers introduced here.

@@

\subsection{Cosmic infrared Background (CIB)}
\label{sec:cib}

@@

Following \cite{Addison:2011se}, we model the CIB power directly as $C_{\ell}^{{\rm CIB}\mbox{-}{\rm CIB}} = 3500 (l/3000)^{-1.25} {\rm Jy^2 / sr}$.
This model provides an accurate fit to several experimental results.
For the cross-spectra with the CMB lensing or other galaxy tracers, $C_{\ell}^{{\rm CIB}\mbox{-}j}$, we use the single-SED model of~\cite{hall:2010}.
It corresponds to the kernel:
\begin{align}\label{eq:12}
W^{\text{CIB}}(z) = b_c\ \frac{\chi^2(z)}{H(z)(1+z)^2}\ e^{-\frac{(z-z_c)^2}{2\sigma^2_z} } f_{\nu(1+z)},
\end {align}
for
\begin{equation}
f_{\nu} = 
\begin{cases}
\Big( e^{\frac{h\nu}{kT}} - 1 \Big)^{-1} \nu^{\beta+3} & (\nu \leq v^{\prime}) \\ \Big( e^{\frac{h\nu^{\prime}}{kT}} - 1 \Big)^{-1} \nu^{\prime \beta+3} \Big( \frac{\nu}{\nu^{\prime}} \Big)^{-\alpha} & (\nu > v^{\prime})
\end{cases}
\end{equation}

We place the peak of the CIB emissivity at redshift $z_c = 2$ with a broad redshift kernel of width $\sigma_z = 2$ and we set $T$ = 34K and $\nu^{\prime} \approx$ 4955 GHz.
\reffig{kernel} shows how the CIB kernel peaks at higher redshift compared to other galaxy survey kernels, with a better overlap with the CMB lensing one.

There are several available CIB observations that have already been used to delens the CMB. Given its full-sky coverage, a promising one is the CIB map derived in \cite{2016A&A...596A.109P} using multi-frequency Planck data and the Generalized Needlet Internal Linear Combination (GNILC) component separation algorithm.

\chapter{Gravitational lensing B-mode and delensing}
\label{sec:th}
The large scale structures described in \refsec{model} have an important impact on the search of primordial CMB B-modes: they lens the primordial E-modes generating non inflationary B-modes that, then constitute an important source of noise.

Indeed the Q and U mode decompositions of the CMB photons polarization are remapped by lensing as:
\be
Q(\hat{\mathbf{n}}) = Q_{\mathrm{unlensed}}(\hat{\mathbf{n}}+\nabla\phi);~~
U(\hat{\mathbf{n}}) = U_{\mathrm{unlensed}}(\hat{\mathbf{n}}+\nabla\phi)
\ee
where the deflection angle is the gradient of the lensing potential integrated along the line of sight $\nabla\phi$.
The CMB polarization is usually decomposed into odd-parity Fourier modes E and B. 
As shown in \cite{seljak:1997}, because of the symmetry of the problem, tensor perturbations are the principal source of the B-modes configuration. For this reason, B-modes are a promising signature of early universe tensor perturbations.

However, this promising sign of primordial gravitational waves is partially obscured by gravitational interactions with large scale structures that generate CMB B-modes by distorting primordial E-modes.
At first order, given the convergence field $\kappa= -\frac{1}{2}\nabla^2\phi$ introduced in \refsec{kappaCMB} the B-modes resulting from the lensing of the primordial E-modes are:
\be
B^{\mathrm{lens}}(\bl) =  \int \frac{d^2 \bl'}{(2 \pi)^2} W(\bl,\bl') E(\ublu) \kappa(\bl - \bl')
\label{eqn:blens}
\ee
where different modes contributes with a different weight:
\be
W(\bl,\bl') = \frac{2 \bl' \cdot (\bl-\bl')}{|\bl-\bl'|^2} \sin(2\varphi_{\bl,\bl'}).
\ee
Here $\varphi_{\bl,\bl'}$ is the angle between the two different modes $\bl$ and $\bl'$.
From this we get the power spectrum of the lensing component of the B-modes:
\be
C_{\ell}^{BB,\mathrm{lens}}  = \int \frac{d^2 \bl'}{(2 \pi)^2} W^2 (\bl,\bl')C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}.
\ee

The B-mode power spectrum measured on the sky is composed of a possible primordial component $C_{\ell}^{BB,r}$ together with the lensing $C_{\ell}^{BB,\mathrm{lens}}$ contribution and the instrumental noise $N_{\ell}^{BB}$ (defined in \refeq{noise}):
\be
C_{\ell}^{BB,\mathrm{measured}} = C_{\ell}^{BB,r} + C_{\ell}^{BB,\mathrm{lens}} + N_{\ell}^{BB}.
\ee
The lensing component is a significant source of B-modes that, at large scales, corresponds to a white noise source of roughly $5 \mu K$-arcmin independent of the angular scale. This means that it is not only larger than the allowed inflationary component at scales smaller than several degrees, but it is also comparable to current levels of instrumental noise. 
For this reason, it is critical to characterize and eventually remove it from the data.
To do so, we build a template \refeq{blens} of the lensing B-modes in the observed patch given a measurement of the E-mode field and the lensing potential $\phi$.
While E is measured directly, we can estimate $\phi$ using "tracers" of the matter distribution that sources the potential.

\subsection{Single tracer of the lensing potential}\label{sec:single-tracer}

We will now show how the delensing efficiency is related to the fidelity of the lensing tracers and the instrumental noise in the CMB E-modes.
If we have a large scale structure field I($\nver$) that traces the lensing potential we can build a template of the lensing B-modes on the sky by a weighted convolution:
\be
\hat{B}^{\mathrm{lens}}(\bl) = \int \frac{d^2 \bl'}{(2 \pi)^2} W(\bl,\bl') f(\bl,\bl') E^N(\ublu) I(\bl - \bl'),
\ee
where $f(\bl,\bl')$ can be determined by minimizing the difference with the true $B^{\mathrm{lens}}(\bl)$ defined in \refeq{blens}. We include the instrumental noise in the CMB E-modes ($E^N$) that will also limit the ability to fully reconstruct the lensing B-modes.

@@

The optimal weights $f(\bl,\bl')$, chosen such that the residual lensing B mode power is minimized, are:
\be
\label{eqn:fweight}
f(\bl,\bl') = \left(\frac{C^{EE}_{{l'}}}{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right)  \frac{C^{\kappa I}_{|\bl-\bl'|}}{C^{II}_{|\bl-\bl'|} }.
\ee
Here $C^{\kappa I}$ and $C^{II}$ are the cross-correlation spectrum of the tracer $I$ with the lensing convergence $\kappa$ and its autospectrum; they are described for each LSS field in \refsec{model}. The power spectrum of the E-modes noise $N^{EE}$ is the same as the B-modes one in \refeq{noise}.

With this choice of $f(\bl,\bl')$ we find that the residual power is:
\beqn
\label{eqn:Bres}
C_{{l}}^{BB,\mathrm{res}} &=&   \int \frac{d^2 \bl'}{(2 \pi)^2}  W^2 (\bl,\bl')
 C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}  \\
&\times&  \left[1 - \left(\frac{C^{EE}_{{l'}} }{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right) \rho^2_{|\bl-\bl'|} \right] \nonumber
\eeqn
with
\be\label{eqn:rho1}
\rho_{\ell}^{2}= \frac{(\cl^{\kappa I})^{2}}{\cl^{\kappa \kappa}\cl^{I I}}.
\ee
